---
title: "Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century marked the beginning of an era characterized by global trade, exploration, and colonization, which often led to profound impacts on indigenous societies in many parts of the world. The forms of European colonization and power relations varied across regions, depending on specific spatial and temporal contexts [@Jordan2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology, as it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was primarily driven by the procurement of Asian luxury goods, raw materials, or natural resources based on commercial interest [@Prakash1997]. The new exchange systems and governance policies imposed by European colonizers often had profound effects on local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, Indigenous communities did not just passively accept European colonization. Recent discussions on colonial interactions also highlight the diverse interactions employed by Indigenous communities in response to European presence, such as resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998]. 

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial center were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Northeastern Taiwan, specifically Yilan, serves as an example of a peripheral case in the 17th century when Europeans established their colonial base in northern Taiwan for a better access to Asian market. Due to its geologically challenging and relatively inaccessible location, Yilan experienced limited direct European control. However, local settlements were frequently involved in the regional exchange network stimulated by the colonial presence in the north. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export in a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017]. Despite some discussions on the policies and operations of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people to the commercial activities are rare. The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jorden2009]. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), a largest settlement in northeastern Taiwan [@Chen2007]. Kiwulan experienced periods of European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. I will primarily focus on subsistence practices, as the choices made in these practices can reveal the strategies Indigenous people employed to cope with European colonial activities and the associated changes in their environment. My research questions are: What were the subsistence practices of Indigenous people before and after the European presence? Can we identify shifts in these strategies in response to colonial economic demands? To address these questions, I will examine the faunal and plant remains from Kiwulan, alongside a preliminary analysis of ceramic residues for investigating culinary practices. Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

# European encounters in the northern region of Taiwan
<!-- the Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Despite the intention to establish a colonial city, San Salvador remained in its early developmental stages, primarily consisting of a fortress and a church, with a population of Spanish officers, soldiers, and missionaries [@Borao2009, pp. 103-134; @Berrocal2018]. The area neighboring San Salvador was inhabited by the Taparris and Quimaurris, local indigenous groups who had resided at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. During Spanish colonization (1626-1642), the colonial processes may not have involved extensive colonialism or asymmetrical power dynamics, with territorial control likely limited to the fort. Written records mention that the Spanish appeared to rely on the local indigenous people for exchanging local goods and food supplies [@Borao2001, SIT:170-171]. This is further supported by archaeological evidence at San Salvador [@Berrocal2018], where only a few European materials were found, alongside more indigenous and Chinese items that might have been used by the Europeans, indicating cultural entanglement.

<!-- the Dutch domination and the processes by which it was established, how and why colonial powers were able to exert control -->
In 1642, the Dutch defeated the Spanish and took over the San Salvador and the Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662. Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007]. For example, the VOC governors used various colonial institutions to secure their authority, including annual landdag gathering, taxation, tribute system, village lease system, census investigations, and the selection of indigenous representatives [@Kang2016; @Andrade1997; @Andrade2007]. Among the political institutions, the tribute system operated from 1642 to 1647 and village lease system introduced in 1644 had great impacts on local indigenous economy and environment. Based on tribute system, each household of allied villages was required to pay rice or high-quality deerskin annually. The amount of deerskins collected from indigenous people in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007]. Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on indigenous lands during certain periods [@Heyns2001]. From 1633 to 1661, the average of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017]. Those colonial activities accelerated the decline in the deer population and could lead to a harsh situation for the indigenous economy .

<!--the variation of colonial powers over native populations, indirect colonization at Kavalan-->
Although Dutch power expanded across nearly all of Taiwan in the mid-17th century, their colonial control and domination varied significantly in different geographical regions. In areas distant from the European colonial bases, colonial control was often challenging and largely unsuccessful. Yilan in northeastern Taiwan is an example of minimal European presence and indirect colonization, with fewer imposed colonial rules for most of the time [@Kang2012]. Yilan is an alluvial plain located about 70 kilometers from the European colony in Keelung, separated by mountains, and inhabited mainly by the local indigenous group, the Kavalan. Given its strategic location on the route to Torrobouan, where the Dutch believed gold mines existed, they were eager to bring this area under their control [@DZI2000; @Kang2016, p47]. In 1644, the Dutch invaded Yilan to incorporate the defeated Kavalan villages into their northern colonial territory, and demanded tribute from the indigenous people [@Kang2016]. nstead of engaging directly with the Kavalan people, the Europeans relied heavily on the assistance of the indigenous Basay people from Keelung to coordinate and collect tributes in Yilan [@Kang2016]. This indirect approach allowed the maintenance of autonomy and flexibility of Kavalan people for negotiation tributes, such as deerskins or products, requested by the VOC [@Kang2016]. 

<!-- pre-exiting trade network -->
Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the early 15th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005]. Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the local Basay people [@Chen2005; @Andrade2007]. This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003]. This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005]. Yilan experienced frequent trading activities driven by the pursuit of foreign goods even before the Europeans arrival, and these activities continued during the European presence in northern Taiwan, likely with some occasional pressure from European powers. This unqiue historical and spatial contexts make Yilan an ideal location to study indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp. 31-78; @Trabert2017].

# Kiwulan site and previous studies
This study focuses on archaeological evidence from Kiwulan, a large multi-component site where an upper component spanning from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. Kiwulan was inhabited by the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in the historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995]. 

The salvage excavation at Kiwulan uncovered 262 units (4 x 4 m) and revealed abundant artifacts and features, such as middens, burials, and post-holes with posts in situ. These findings suggest a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. In addition to locally made pottery with geometric patterns, foreign goods such as porcelains, stonewares, glass beads, carnelian beads and metal artifacts were largely found in the residential area and burials. Since there is no evidence of local production of beads or metallurgy, these goods are believed to have been acquired through a regional exchange network with people from outside the area, mostly with the Basay people active in northern Taiwan. 

# Previous research of 

# Mateirals and methods

# Results

<!-- Here's some example analysis code: -->

```{r}
#| label: get-data
#| eval: false
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

```{r}
#| label: fig-demo-plot
#| fig-cap: "A plot of random numbers"
plot(rnorm(10))
```

@fig-demo-plot shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`

```{r}
#| label: demo-inline-code
#| echo: false
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion
Taiwan has played an important role in exploring the effect of colonial activities on Indigenous communities and local environment since the protohistoric period. This project will explore the environmental effect of colonial presence on local ecology on a regional scale using archaeological evidence that has not yet been fully understood. Archaeological materials can offer more direct evidence, allowing better interpretation for the environments where Indigenous live in that can also reduce biases from historical records. Through exploring the long-term dynamics of human-environmental relationship in colonial settings, we can better understand changes in local environments and Indigenous resilience. In addition to the Europeans, the interaction between Chinese traders and Indigenous groups in Taiwan since the Neolithic period, and an influx of Han Chinese immigrants in the 18th century, has contributed to the significant local cultural and environmental transformation. Those historical events are closely related to the Taiwan Indigenous people today, which is crucial to understanding the contemporary Indigenous cultures and socio-ecological landscape. 
The results of this project can advance our knowledge of the reaction of Indigenous people in Taiwan when facing dominant foreign power, and generate new anthropological insights about the adaptation and choices of Indigenous people. In the long term, this research will also benefit local Indigenous people, Kavalan, in northeastern Taiwan and help them to trace back their past and revitalize their cultural heritage by the plans of public presentation in the future. In addition, this project will combine geoarchaeological, geochemical, and computational methods to analyze archaeological evidence. The multidisciplinary approach can not only advance the development and application of scientific techniques in Taiwan archaeology, but also provide an opportunity to form collaborative relationships with scholars with different backgrounds, which will help to develop the academic network and create interdisciplinary collaboration.

This research will contribute to a broader understanding of the diverse ways Indigenous communities adapted to and resisted colonial forces, enhancing our comprehension of the long-term impacts of European colonization on local societies.

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
