---
title: "Exploring environmental impacts of colonial contacts on Indigenous communities in northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The actual document text starts here: -->

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century was usually associated with long-distance trade, exploration, and colonization in many parts of the world, which often led to profound impacts on indigenous societies. Depending on particular geographic and historical circumstances, many regions experienced different forms of European colonialism and power relations [@Jordan2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology because it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In Asia, European colonization was mainly motivated by the acquisition of raw materials, natural resources, or luxury items from Asia for commercial purposes [@Prakash1997]. Although the new exchange systems or colonial institutions introduced by Europeans could change local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002], indigenous groups did not just accept European colonization passively. Recent discussions on colonial interactions also suggest the active role of indigenous groups in responding to the arrival of Europeans and presenting resistance, negotiation, persistence, resilience, and entanglement [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998]. 

<!-- NE Taiwan as a case study -->
While many studies focus on prominent colonies, places on the periphery of colonial centers were also more and less impacted by European powers or associated trading systems [@Acabado2017; @Trabert2017]. Exploring these areas is important to gain a broader understanding of the varied consequences of colonial expansion. Yilan, located in northeastern Taiwan, is an example of a periphery region where European control was limited and indirect for most of the time but intensively involved in regional trade activities stimulated by the European presence. In the mid-17th century, the VOC (Dutch East India Company) formalized the deerskin trade in Taiwan for export on a large scale, involving many Indigenous communities and Han Chinese settlers [@Heyns2001; @Cheng2017]. Despite some discussions on the operations and policies of the deerskin trade, direct investigations on archaeological evidence into the experiences of local indigenous people with the commercial activities are rare. The European impacts on local ecological conditions have been recorded in many places in colonial settings [@Jordan2009]. Whether the interaction between Indigenous people and their environment changed due to European trade demands requires further archaeological investigation.

<!-- research aim and questions -->
In this paper, I will explore the interactions between people and their environment in a colonial context by examining the resource use of indigenous people through a case study at the Kiwulan site (1350-1950 CE), the largest Iron Age settlement in northeastern Taiwan [@Chen2007]. Kiwulan encountered European presence in the 17th century, followed by a large influx of Han Chinese migration in the 19th century. What was the resource use of Indigenous people before and after the foreign presence? Can we identify shifts in these practices in response to colonial economic demands? To address these questions, I will focus on subsistence practices because the choices made in the practices can reveal the strategies Indigenous people used to respond to European exploitation with associated changes in their environment. I will primarily examine the zooarchaeological assemblages from Kiwulan for subsistence practices, coupled with botanical evidence as a supplementary evidence. Zooarchaeology can offer insights into how animals were utilized in response to economic strategies within certain social contexts, such as intercultural contact or European colonization (DeFrance, 2009). Through an environmental perspective, this study aims to shed light on Indigenous persistence and resilience strategies in the face of foreign interruptions.

# European Encounters in the Northern Taiwan
<!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui, to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The neighboring area of the fort was inhavited by local indigenous groups, the Taparris and the Quimaurris, who had lived at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and indigenous people might not involve extensive colonialism or intercultural domination, and the Spanish territorial control could be limited to the fort. Instead, the Spanish relied on the exchange of local goods and food surpluses with local indigenous people, according to written records [@Borao2001, SIT:170-171]. This corresponds with a higher proportion of indigenous and Chinese items with little European material found in San Salvador [@Berrocal2018], indicating cultural entanglement. 

<!-- the Dutch domination, how and why colonial powers were able to exert control -->
In 1642, the Dutch defeated the Spanish and took over San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662. Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007]. For example, the VOC governors used various colonial institutions to secure their authority, including the annual landdag gathering, taxation, tribute system, village lease system, census investigations, and the selection of indigenous representatives [@Kang2016; @Andrade1997; @Andrade2007]. Among the political institutions, the tribute system operated from 1642 to 1647 and the village lease system introduced in 1644 had great impacts on the local indigenous economy and environment. Based on the tribute system, each household in allied villages was required to pay rice or high-quality deerskin annually. The amount of deerskins collected from indigenous people in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007]. Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on indigenous lands during certain periods [@Heyns2001]. From 1633 to 1661, the average number of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017]. Those colonial activities accelerated the decline in the deer population and could lead to a harsh situation for the indigenous economy.

<!--variation of colonial powers across regions and indirect colonization at Kavalan-->
Although the Dutch power expanded across nearly all of Taiwan in the middle of the 17th century, the extent of their colonial dominance and control differed greatly among different regions. Colonial rules in regions far from the European forts were usually difficult and unsuccessful. Yilan in northeastern Taiwan is an example of minimal European presence and indirect colonization, with fewer imposed colonial rules for most of the time [@Kang2012]. Yilan is an alluvial plain, located about 70 kilometers from the European colony in Keelung, with mountains in between. The main inhabitants in Yilan were the local indigenous group, the Kavalan. Given its strategic location on the route to Torrobouan, where the Dutch believed gold mines existed, they were eager to bring this area under their control [@DZI2000; @Kang2016, p47]. In 1644, the Dutch invaded Yilan to incorporate the defeated Kavalan villages into their northern colonial territory and demanded tribute from the indigenous people [@Kang2016]. Instead of engaging directly with the Kavalan people, the Europeans relied heavily on the assistance of the Basay people active in northern Taiwan to coordinate and collect tributes in Yilan [@Kang2016]. This indirect approach allowed the maintenance of autonomy and flexibility for the Kavalan people, who were able to negotiate tribute requested by the VOC [@Kang2016]. 

<!-- the pre-existing trade network in norhtern Taiwan-->
Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the early 15th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005]. Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the local Basay people [@Chen2005; @Andrade2007]. This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003]. This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005]. Yilan experienced frequent trading activities driven by the pursuit of foreign goods even before the Europeans arrival, and these activities continued during the European presence in northern Taiwan, likely with some occasional pressure from European powers. This unique historical and spatial context makes Yilan an ideal location to study indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp. 31-78; @Trabert2017].

# Archaeological Context at Kiwulan
<!--introduction of Kiwulan and the population-->
Kiwulan, located in the northern part of Yilan, is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed. A total of 262 excavation units uncovered abundant artifacts and features, including middens, burials, and post-holes with wooden posts in situ. It suggests that Kiwulan is a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. The main inhabitants of Kiwulan were the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995]. 

<!--archaeological materials-->
The most common artifact was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan [@TaiwanHistory2005]. In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts, were largely found in both the residential area and burials. Due to no evidence of local production of beads or metallurgy, these items are believed to have been acquired through exchange, primarily with the Basay people active in northern Taiwan. This indicates that Kiwulan has continuously participated in regional trade networks closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@Wang2007]. Moreover, the use of foreign goods appears to have cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].

<!--archaeological remains-->
There were a wide range of animal bones and plant remains uncovered, reflecting the diversity of the local diet and the environment close to river and coastal area. The animals include deer, pigs, cows, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004]. Rice farming seems to be common but limited to a household scale, according to historical records [@]. Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments, which indicates a reliance on aquatic and marine resources. Those archaeological remains correspond with tools related to subsistence found at Kiwulan, including iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities. According to those pieces of evidence, the subsistence practices at Kiwulan should be primarily reliant on rice farming, supplemented by hunting and fishing for wild resources. The resource use during the periods of external impacts still need further examination to better understands the changes over time. 

# Materials and Methods
To explore changes in dietary practices before and after the European arrival with associated long-distance trade, I examined faunal assemblages from 40 adjacent units at the residential area excavated during the first season (2001-2003) at Kiwulan [@Chen2007]. The remains considered in this study are from dwelling spaces and trash pits that could reflect daily dietary practices, whereas those from burials that might relate to ritual activities are excluded. These 40 units were selected because of their location that were believed to be the center of the settlement with thick archaeological deposits and relatively intact archaeological contexts. Also, the spatial continuity can provides a more consistent framework for understanding food practices over time on a settlement scale. The same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences. 

To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six temporal layers spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could offer a relatively robust framework for comparing archaeological records before and after the European influence. Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. Deer antlers that were not attached to cranial bones were excluded since they might not directly related to dietary practices. To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers. Taxonomic abundance was quantified using the Number of Identified Specimens (NISP) method [@Grayson2004; @Lyman2012]. Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size, and therefore is used throughout this study [@Grayson1984; @Lyman2008]. For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the average number of skeletal elements of that given taxa. 

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the skeletal compositions, fragmentation, and butchery patterns of deer bones were examined. The distribution of body portions and butcher patterns can shed light on how indigenous people utilize deer in response to economic demands during the foreign presence. The proportion of deer skeletal elements was quantified using NISP and classified into groups based on the meat yield: meaty elements (e.g., upper limb), mid-meaty elements (e.g., body parts), and non-meaty elements (e.g., skull bones and lower limb) [@Madrigal2004]. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain. Thus, non-meaty elements were considered as secondary source of nutrition in the interpretation. Similarly, the degree of fragmentation and the weight of broken unidentified bones could also imply the extraction of marrow or grease. The degree of fragmentation is calculated by dividing Minimal Number of Elements (MNE) by NISP. For butcher patterns, cut marks on major limb joints, including the shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were identified and quantified for a comparison between different temporal sequences. In addition, modifications such as cut marks, fracturing, and burning were examined, especially the cut marks associated with the skinning process. Skinning processes can be identified by the encircling and straight marks left on the lower limb, such as metapodials, which result from the series of cuts typically made to acquire intact animal hides [@Costamagno2019; @Isaakidou2007].

# Results

```{r setup, include= FALSE}
library(tidyverse)
library(knitr)
```

```{r}
#| label: get-data
#| echo: false
#| cache.extra: tools::md5sum('KWL_faunal_P.xlsx')

# read data
kwl_fauna <- readxl::read_excel(here::here("analysis","data","raw_data","KWL_faunal_P.xlsx"), sheet = 1,
                                col_types = "text")

kwl_fauna_HM <- readxl::read_excel(here::here("analysis","data","raw_data","KWL_faunal_HM.xlsx"), sheet = 1,
                                   col_types = "text")

kwl_fauna_broken <- readxl::read_excel(here::here("analysis","data","raw_data","broken_fauna.xlsx"), sheet = 1,
                                   col_types = "text")

sample_p <- readr::read_csv(here::here("analysis", "data", "raw_data", "kwl-list-of-sampling-squares.csv"))

kwl_chro_6 <- readxl::read_excel(here::here("analysis", "data", "raw_data", "KWL_chronology.xlsx"), sheet = 5)

# tidy chronological data
kwl_chro_tidy_6 <-
  kwl_chro_6 %>%
  mutate(area = str_sub(Pit, 5, 7), Pit = str_sub(Pit, 1, 4)) %>%
  mutate(area = ifelse(area == "", "ABCD", area)) %>%
  mutate(area = sapply(strsplit(area, ""), paste, collapse= ",")) %>%
  separate_rows(area) %>%
  pivot_longer(cols = starts_with("L"), names_to = "layer", values_to = "period") %>%
  mutate(layer = str_sub(layer, 2, 3)) %>%
  mutate(layer = ifelse(nchar(layer) == 1, paste0("0", layer), layer))

# combine faunal data of cultural layers and temporal sequences
fauna_sam <-
  kwl_fauna %>%
  mutate(Pit = paste(context, pit, sep = "")) %>%
  filter(Pit %in% sample_p$the_sq) %>%
  filter(componant == "上文化層" &!layer == "00") %>% #remove sc
  #filter(!str_detect(`總報告類別`, "[[:punct:]]+|[0-9]+")) %>%
  mutate(area = ifelse(is.na(area), "C", area)) %>% # assign an area to the blanks, which won't change the results
  left_join(kwl_chro_tidy_6, by = c("Pit" = "Pit", "layer" = "layer", "area" = "area"))

# combine faunal data of middens and temporal sequences
fauna_H_sam <-
  kwl_fauna_HM %>%
  select(-`空間號`, -`層   位`) %>%
  mutate(Pit = paste("P", pit, sep = "")) %>%
  filter(Pit %in% sample_p$the_sq) %>%
  filter(componant == "上文化層") %>%
  left_join(kwl_chro_tidy_6, by = c("Pit" = "Pit", "layer" = "layer", "area" = "area"))

fauna_combined_context <-
  rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
  mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
  group_by(taxa) %>%
  mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
  ungroup() %>%
  mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
  mutate(category = case_when(str_detect(animal, "e deer") ~ "deer", str_detect(animal, "fish") ~ "fish",
                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
                              animal == "aves" ~ "bird", TRUE ~ animal)) %>%
  mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
                           str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
                           str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
                           str_detect(category, "fish") ~ "fish", str_detect(category, "bird") ~ "bird",
                           TRUE ~ "reptile")) %>%
  mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
  mutate(category = fct_relevel(category, "deer","boar","muntjac","cattle","rat","bird","fish","turtle")) %>%
  mutate(modify = case_when(str_detect(`人為痕跡`, "切")|str_detect(`人為痕跡`, "削")|
                            str_detect(`人為痕跡`, "砍")~ "cutmarks",
                            str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒")~ "burn", TRUE ~ "no")) %>%
  filter(is.na(refitted)) # for NISP
```

```{r}
#| label: tbl-counts
#| echo: false
#| tbl-cap: "Taxa, Number of Identified Specimens (NISP), and weight of faunal remains from the sampled area at Kiwulan."
#| tbl-colwidths: [40,25,15,20]
#| tbl-cap-location: bottom
# taxa counts
fauna_combined_taxa <-
  fauna_combined_context %>%
  count(taxa, `Weight (g)`) %>%
  mutate(taxa = str_replace(taxa, "cervidae", "cervid")) %>%
  mutate(taxa = case_when(str_detect(taxa, "mm") ~ paste("Unidentified", tolower(taxa), sep =" "),
                          TRUE ~ taxa)) %>%
  mutate(`Common name` = case_when(str_detect(taxa, "Cer") ~ "Sika deer", str_detect(taxa, "Bub") ~ 
                                   "Water buffalo", str_detect(taxa, "Rus") ~ "Sambar deer",   
                                   str_detect(taxa, "Ree") ~ "Muntjac", str_detect(taxa, "Sus") ~ "Boar", 
                                   str_detect(taxa, "Bos") ~ "Yellow cattle", str_detect(taxa, "Rat") ~ 
                                   "Rat", str_detect(taxa, "Pli") ~ "Catfish", str_detect(taxa, "Act") ~ 
                                   "Ray-finned fish", str_detect(taxa, "Tes") ~ "Turtle",
                                   str_detect(taxa, "cervid") ~ "Deer family")) %>%
  mutate(group = case_when(str_detect(`Common name`, "deer") ~ 1, str_detect(`Common name`, "Deer") ~ 2,
                           str_detect(taxa, "Sus") ~ 3, str_detect(taxa, "Bub")|str_detect(taxa, "Bos") ~ 4,
                           str_detect(taxa, "munt") ~ 5, str_detect(taxa, "Rat") ~ 6, 
                           str_detect(`Common name`, "fish") ~ 7, str_detect(taxa, "Tes") ~ 8,
                           str_detect(taxa, "Ave") ~ 9, str_detect(taxa, "large m") ~ 10,
                           str_detect(taxa, "medium m") ~ 11, TRUE ~ 12)) %>%
  group_by(group) %>%
  arrange(desc(`Weight (g)`), .by_group = TRUE) %>%
  ungroup () %>%
  rename(Taxon = taxa, NISP = n) %>%
  select(Taxon, `Common name`,  NISP, `Weight (g)`) %>%
  bind_rows(summarise(., across(where(is.numeric), sum), across(where(is.character), ~"Total"))) %>% 
  mutate(`Common name`= ifelse(is.na(`Common name`)|`Common name` == "Total", "-", `Common name`))

total_NISP <- fauna_combined_taxa %>% select(NISP) %>% tail(1) #rev(fauna_combined_taxa$NISP)[1]
unident <- fauna_combined_taxa %>% filter(str_detect(Taxon, "Uni")) %>% summarise(sum(NISP))
deer_NISP <- fauna_combined_taxa %>% 
  filter(str_detect(`Common name`, "eer")|str_detect(Taxon, "Ree")) %>% summarise(sum(NISP))
deer_NISP_per <- round(deer_NISP/(total_NISP-unident), 1)*100


library(kableExtra)
knitr::kable(fauna_combined_taxa, align = "l") 
```

The NISP of faunal assemblages from Kiwulan analyzed in this study is `r total_NISP`, where deer family is the majority that accounts for about `r deer_NISP_per`% of the identified bones, followed by boar and cattle as the second and third source based on NISP (@tbl-counts). Within the bones from deer family, sika deer's bone is the most common one, with a few bones from muntjact and a specimen indicating a possible affiliation with sambar deer. Although sika deer was the major animal source at the site, variations could be observed between different cultural layers across 600 years. To examine and compare general patterns of animal use in terms of their economic values over time, deer family were separated into two categories, deer (sika deer, sambar deer, and large deer) and muntjac, while water buffalo and yellow cattle (NISP = 1) were combined into the cattle category. 


```{r}
#| label: ubiquity
#| echo: false
#| tbl-cap: "Ubiquity of faunal categories over the 40 units"
#| tbl-colwidths: [60,40]
# calculate the number of units for each cultural layers
kwl_unit_per_period <-
  kwl_chro_6 %>%
  mutate(all = pmap_chr(select(., -Pit, -Grid), ~toString(unique(na.omit(c(...)))))) %>%  # unite and remove duplicates
  mutate(unit = str_extract(Pit, "[A-Z]+[0-9]{3}")) %>%
  distinct(unit, .keep_all = TRUE) %>%
  mutate(CL1 = ifelse(str_detect(all, "CL1"), unit, NA)) %>%
  mutate(CL2 = ifelse(str_detect(all, "CL2"), unit, NA)) %>%
  mutate(CL3 = ifelse(str_detect(all, "CL3"), unit, NA)) %>%
  mutate(CL4 = ifelse(str_detect(all, "CL4"), unit, NA)) %>%
  mutate(CL5 = ifelse(str_detect(all, "CL5"), unit, NA)) %>%
  mutate(CL6 = ifelse(str_detect(all, "CL6"), unit, NA))

# count the units by cultural layers for later join
unit_by_period <-
  data.frame(unit_count = colSums(!is.na(kwl_unit_per_period))) %>%
  slice_tail(n = 6) %>%
  tibble::rownames_to_column("period")

# ubiquity of taxa from the 40 units by temporal sequences
fauna_ubiquity <-
  fauna_combined_context %>%
  select(Pit, layer, category, `重量(g)`, period, taxa) %>%
  count(period, Pit, category) %>%
  count(period, category) %>%
  drop_na(period, category) %>%
  left_join(unit_by_period) %>% # join the unit counts for each period
  mutate(ubiquity = paste(round(n/unit_count *100, 2), "%")) %>%
  select(-n, -unit_count) %>%
  pivot_wider(names_from = period, values_from = ubiquity) %>%
  mutate(across(CL1:CL6, ~replace_na(.x, "-"))) %>% 
  arrange(category)

knitr::kable(fauna_ubiquity)
```

@ubiquity shows how we can have a caption and cross-reference for a plot. Note that figure label and cross-references must both be prefixed with `fig-`

```{r}
#| label: demo-inline-code
#| echo: false
x <- round(pi, 2)
```

Here is an example of inline code `r x` in the middle of a sentence.

# Discussion

# Conclusion
Taiwan has played an important role in exploring the effect of colonial activities on Indigenous communities and local environment since the protohistoric period. This project will explore the environmental effect of colonial presence on local ecology on a regional scale using archaeological evidence that has not yet been fully understood. Archaeological materials can offer more direct evidence, allowing better interpretation for the environments where Indigenous live in that can also reduce biases from historical records. Through exploring the long-term dynamics of human-environmental relationship in colonial settings, we can better understand changes in local environments and Indigenous resilience. In addition to the Europeans, the interaction between Chinese traders and Indigenous groups in Taiwan since the Neolithic period, and an influx of Han Chinese immigrants in the 18th century, has contributed to the significant local cultural and environmental transformation. Those historical events are closely related to the Taiwan Indigenous people today, which is crucial to understanding the contemporary Indigenous cultures and socio-ecological landscape. 
The results of this project can advance our knowledge of the reaction of Indigenous people in Taiwan when facing dominant foreign power, and generate new anthropological insights about the adaptation and choices of Indigenous people. In the long term, this research will also benefit local Indigenous people, Kavalan, in northeastern Taiwan and help them to trace back their past and revitalize their cultural heritage by the plans of public presentation in the future. In addition, this project will combine geoarchaeological, geochemical, and computational methods to analyze archaeological evidence. The multidisciplinary approach can not only advance the development and application of scientific techniques in Taiwan archaeology, but also provide an opportunity to form collaborative relationships with scholars with different backgrounds, which will help to develop the academic network and create interdisciplinary collaboration.

This research will contribute to a broader understanding of the diverse ways Indigenous communities adapted to and resisted colonial forces, enhancing our comprehension of the long-term impacts of European colonization on local societies.

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

::: {#refs}
:::

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r}
#| label: colophon
#| cache: false

# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```
